---
title: "pcvr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pcvr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# devtools::install_github("joshqsumner/pcvr")
library(pcvr)
library(ggplot2)
```

## Example workflow for a relatively small dataset

This workflow will use example data from github to analyze single value traits such as area, perimeter, or height and multi-value traits such as Fv/Fm or Chlorophyll Index (ci_rededge). 

- read data from pcvrTestData online repository using read.pcv
- download pcvrTestData and read from local using read.pcv with filters
- use pcv.boxplot for something, show comparisons
- use pcv.joyplot for multi-value traits

First we read in our data in wide format, keeping only the single value traits.

```{r}
df<-read.pcv("https://raw.githubusercontent.com/joshqsumner/pcvrTestData/main/pcvrTest2.csv", mode = "wide", singleValueOnly = T)
```

This dataset is small enough that it is no problem to load the entire data and store it in memory. For larger data that may be prohibitive in which case the other features of `read.pcv` should be used.

```{r, eval=F}
download.file(url="https://raw.githubusercontent.com/joshqsumner/pcvrTestData/main/pcvrTest2.csv", destfile="example.csv")
ex<-read.pcv("example.csv", mode = "wide", reader="read.csv", filters = "trait in area, perimeter")
```

Now we want to look into some of our single value phenotypes. `pcvBox` is a simple wrapper around ggplot2 functions. Use it or don't.

```{r}
pcvBox(df, x="timepoint", y="area.pixels", fill="genotype")
```

`pcvBox` also uses ggpubr to do difference of means testing (currently with Welch's T tests) with a slightly simplified interface for common tasks.

```{r}
pcvBox(df[df$timepoint=="tp4",], x="genotype", y="area.pixels", fill="genotype", compare="G1") # compare one level to all others
pcvBox(df[df$timepoint=="tp4",], x="genotype", y="area.pixels", fill="genotype", compare="G2")
pcvBox(df[df$timepoint=="tp4",], x="genotype", y="area.pixels", fill="genotype", compare=c("G3","G6")) # compare a pair of values
pcvBox(df[df$timepoint=="tp4",], x="genotype", y="area.pixels", fill="genotype", compare=list(c("G3","G6"), c("G1","G2"))) # compare a list of pairs.
```

We might also want to compare multi value traits. For that we can read the data in long format and keep those traits.

```{r}
lng<-read.pcv("https://raw.githubusercontent.com/joshqsumner/pcvrTestData/main/pcvrTest2.csv", mode = "long", singleValueOnly = F)
```

Joyplots using ggridges are easily made with `pcv.joyplot`.

```{r}
x<-pcv.joyplot(lng, index = "index_frequencies_index_ndvi", group=c("genotype", "timepoint"))
```

Optionally there are several methods for statistical comparison which makes `pcv.joyplot` return a list with a plot and a dataframe.

```{r}
x<-pcv.joyplot(lng, index = "index_frequencies_index_ndvi", group=c("genotype", "timepoint"), method="ks")
x[["plot"]]
head(x[["stats"]])
```

```{r}
x<-pcv.joyplot(lng, index = "index_frequencies_index_ndvi", group=c("genotype", "timepoint"), method="beta")
x[["plot"]]
head(x[["stats"]])
```





