---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# `pcvr`: An R Package for Plant Phenotyping Statistics

<!-- badges: start -->
  [![R-CMD-check](https://github.com/danforthcenter/pcvr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/danforthcenter/pcvr/actions/workflows/R-CMD-check.yaml)
  [![Coverage
Status](https://codecov.io/github/danforthcenter/pcvr/coverage.svg?branch=master)](https://app.codecov.io/github/danforthcenter/pcvr)
<!-- badges: end -->

`pcvr` provides R functions for use with `PlantCV` output or other phenotype data with the goal of lowering the barrier to entry for Bayesian statistics and non-linear modeling.

## Installation

The release version of `pcvr` can be installed from `CRAN`

```{r, eval = FALSE}
install.packages("pcvr")
library(pcvr)
```

Alternatively the development version of `pcvr` can be installed using remotes/devtools `install_github` as shown below.
Note that the default behavior in devtools/remotes is to only install true dependencies. Some functions in pcvr use specific packages that would otherwise not be needed for most work, notably the `brms` modeling functions. To install suggested packages (see DESCRIPTION file) add `dependencies=T` to the `install_github` function call.

```{r, eval = FALSE}
devtools::install_github("danforthcenter/pcvr")
library(pcvr)
```

## Vignettes

See the `Vignettes` tab above for several example workflows for common plant phenotyping tasks.

## Tutorials

See the `Quarto Tutorials` tab above for links to Quarto presentations on github that go more in depth about the rationale behind several `pcvr` functions/options.

## Function Reference

Functions are separated by broad goal/type of data they use under the `Functions` tab above.

## Getting started

Please see the `bellwether` vignette (named for the high throughput phenotyping facility at the Donald Danforth Plant Science Center) for a general introduction to `pcvr`.

```{r, eval = FALSE}
vignette("bellwether", package="pcvr")
# or 
browseVignettes("pcvr")
```

## Feedback

Please report bugs and make feature requests with issues the [github page](https://github.com/danforthcenter/pcvr).
